---
title: 'Howtos'
output: html_document
#output: rmarkdown::html_vignette
#vignette: >
#  %\VignetteIndexEntry{Howtos}
#  %\VignetteEngine{knitr::rmarkdown}
#  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, message=FALSE, eval=TRUE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, eval=FALSE)
require(ubiquity)
require(deSolve)
require(ggplot2)
require(foreach)
require(doParallel)
require(rhandsontable)
```

## Working with `rxode2`, `nlmixr2`, NONMEM, and Monolix

Say you have a model in ubiquity and you want to use it with nlmixr2. Or perhaps you want to try out the simulation engine of `rxode2`, or you want to hand the model off to someone using `Monolix`. This should provide you with a way to easily convert your model into different formats in an automated fashion. 

### Getting your model in `rxode2` format

As an example consider the system file for the two compartment model  (`system_2cmt.txt` below). When you build the system an `rxode2` output target will be created. You can use the `system_fetch_template()` function to access this file. Below I'm placing the script into the temporary directory. This is a script that will generate an `rxode2` model function and store the function in the object `my_model`. This is the default input for `nlmixr2`. You can then use this function and the piping methodology to do things like add IIV terms to parameters. So you can use `ubiquity` to build your structural model and then piping, `rxode2`, and `nlmixr2` for analysis. 

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
sf_2cmt_txt = "
# Two compartment model with absorption compartment (At). The central
# compartment (Cc) had a volume (Vc), and the tissue compartment (Cp) has a
# volume (Vt). The system is parameterized in terms of macro constants (Q and
# CL)
#
# While the the default dosing is a 1 mg IV dose into the central compartment,
# the model is written to accept dosing into an absorption compartment and
# through continuous IV infusion.
#       _________ 
#      |         |
#      |   At    |
#      |         |
#      |_________|
#           |
#           | ka, fb
#           |
#           V
#       _________         _________  
#      |         |   Q   |         | 
#      |   Cc    |------>|   Cp    |
#      |   Vc    |<------|   Vt    |
#      |_________|       |_________|
#           |
#           | CL
#           |
#           V  
#
# System Units:
#                
#   mass          [=] mg 
#   volume        [=] ml
#   concentration [=] mg/ml
#   time          [=] hr


# #-------------#
# | Parameters  |
# #-------------#
#
# System parameters
#    name              value     lower      upper   units editable    grouping
#                                bound      bound

<P>  Vc                  1.0       eps        Inf   ml         yes     System
<P>  Vt                  1.0       eps        Inf   ml         yes     System
<P>  CL                  1.0       eps        Inf   ml/hr      yes     System
<P>  Q                   1.0       eps        Inf   ml/hr      yes     System
<P>  ka                  1.0       eps        Inf    1/hr      yes     System
<P>  fb                  1.0       eps        Inf   --         yes     System


# Bolus Events
# ------------
# times/events state   values        scale      units
<B:times>;             [  0  ];      1;          hours
<B:events>;      Cc;   [1.0  ];      1/Vc;       mg     
<B:events>;      At;   [0.0  ];      1;          mg     

# Infusion Rates 
# ------------
#  name     time/levels  values  scale    units
<R:Dinf>;     times;       [0];     1;    hours
<R:Dinf>;     levels;      [0];     1;    mg/hour

<ODE:At> -ka*At           
<ODE:Cc>  ka*At*fb/Vc - CL/Vc*Cc -Q*(Cc-Cp)/Vc  + Dinf/Vc
<ODE:Cp>                         +Q*(Cc-Cp)/Vt
 
<O> Cc_mg_ml    = Cc
<VP> prop_err   0.1            eps    inf      --     yes      Variance
<VP> add_err    0.1            eps    inf      ng/ml  yes      Variance
<OE:Cc_mg_ml> add=add_err; prop=prop_err

<TS:hours> 1.0
<TS:days>  1.0/24"
sf_2cmt = file.path("system_2cmt.txt")
sf_2cmt_full = file.path(tempdir(), sf_2cmt)

fileConn = file(sf_2cmt_full)
writeLines(sf_2cmt_txt, fileConn)
close(fileConn)

```

```{r eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
cfg = build_system(system_file         = sf_2cmt_full, 
                   temporary_directory = file.path(tempdir(), "transient"))

fr = system_fetch_template(cfg, 
                           template         = "nlmixr2", 
                           output_directory = tempdir(),
                           overwrite       =  TRUE)

library(rxode2)
source(file.path(tempdir(), "system_nlmixr2.R"))
```

You can see the contents of `system_nlmixr2.R` and what the `my_model` object looks like below. Note that to include the output with `ubiquity` you will need to have some error model. So along with the `<O>` Descriptor you will need to define some variance parameters (`<VP>`) and also an output error model (`<OE:?>`).

### Getting your model in NONMEM format

Once you have your model in `rxode2` format you can convert it further into NONMEM format with the `babelmixr2` package (via the `ruminate` package). To do this you need at least one inter-individual variability term. You can do this in the system file with the `<IIV:?>` delimiter. In this example we will do that by using model piping. Here we specify that the typical value of `Vc`, the parameter `TV_Vc`, should have an error that is lognormally distributed. This will add the IIV term `eta_Vc`.

```{r eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results="hide"}
my_model = my_model |> 
  model({Vc = exp(TV_Vc + eta_Vc)}) 
```

Next we just need to load the `ruminate` package (>=0.2.4) and use the `rx2other()` function to create a NONMEM control stream: 

```{r eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(ruminate)
my_model_nm = rx2other(my_model, out_type="nonmem")
```

### Getting your model in Monolix format

You can also get the model in Monolix by specifying `monolix` as the output type:

```{r eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
my_model_mlx = rx2other(my_model, out_type="monolix")
```

### Further reading

For more discussion on this topic see the following:

* [https://github.com/nlmixr2/babelmixr2/issues/107](https://github.com/nlmixr2/babelmixr2/issues/107)


### Models in different formats {.tabset}

#### `ubiquity`

Contents of `system_2cmt.txt` file:

```{r echo=FALSE, comment='', message=TRUE, eval=TRUE}
cat(readLines(sf_2cmt_full), sep="\n")
```

#### `rxode2` script

Contents of `system_nlmixr2.R` file produced by `system_fetch_template()`:

```{r echo=FALSE, comment='', message=TRUE, eval=TRUE}

cat(readLines( file.path(tempdir(), "system_nlmixr2.R")) , sep="\n")
```

#### `rxode2` object

The `my_model` object:

```{r echo=FALSE, comment='', message=TRUE, eval=TRUE}
my_model
```

#### NONMEM

The file containing the control stream is found here: 

`my_model_nm$files$ctl$fn_full`.

```{r echo=FALSE, comment='', message=TRUE, eval=TRUE}
cat(readLines(  my_model_nm$files$ctl$fn_full ) , sep="\n")
```

#### Monolix

The file containing the mlxtran output is found here: 

`my_model_mlx$files$mlxtran$fn_full`

```{r echo=FALSE, comment='', message=TRUE, eval=TRUE}
cat(readLines(  my_model_mlx$files$mlxtran$fn_full ) , sep="\n")
```