---
title: 'Individual and Population Simulations in Ubiquity'
author: "John Harrold <john.m.harrold@gmail.com>"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ubiquity)
require(deSolve)
require(ggplot2)
require(foreach)
require(doParallel)
require(doRNG)
require(rhandsontable)
```

## Introduction

To demonstrate running simulations we'll use a PK model of mAbs in people
([Davda etal. mAbs, 6(4), 1094-1102](http://doi.org/10.4161/mabs.29095)). You can copy this model to the local folder using ``system_new`` specifying the "mab_pk" system file. The contents of this system file can be seen at the bottom. Next it is necessary to build the system file creating the ubiquity model object (``cfg``). 

```{r results="hide", message=FALSE}
library(ubiquity)
system_new(file_name="system.txt", system_file="mab_pk", overwrite = TRUE)
cfg = build_system(system_file = "system.txt")
```

After building any system you can then create templates. To create a template script for running simulations use ``system_fetch_template`` and specify  ``"Simulation"`` as the template argument:

```{r eval=FALSE}
system_fetch_template  <- function(cfg, template="Simulation")
```

This will create the file ``analysis_simulate.R`` in the working directory. This template should have common options and dosing information commented out. You simply need to uncomment them and run the simulation. 

## Simulating an Individual Response

The system parameters for the default parameter set can be pulled out of this object:

```{r results="hide", warning=FALSE}
cfg = build_system(system_file = "system.txt")
parameters = system_fetch_parameters(cfg)
```

Next different simulation options can be set. For example the following will set the duration of the simulation to three months:

```{r results="hide"}
cfg = system_set_option(cfg, group  = "simulation", 
                             option = "output_times", 
                             seq(0,7*4*3,1))
```

The system file is written to allow both IV (``Cp``) and SC (``At``) dosing. So we zero out any default dosing specified in the system file and specify four SC doses given two weeks apart:
```{r results="hide"}
cfg = system_zero_inputs(cfg)
cfg = system_set_bolus(cfg, state  = "At", 
                            times  = c(  0,  14,  28,  42 ), #  day
                            values = c(200, 200, 200, 200 )) #  mg
```

Next we run the simulation:
```{r results="hide"}
som = run_simulation_ubiquity(parameters, cfg)
```

The variable ``som`` is a list contains the mapped simulation output, and the time course is stored in the ``simout`` element. The first column (``time``) contains the simulation time in the units of the simulation, days in this case. Next there is a column for each state (``At``, ``Cc``, ``Cp``) and a column for each output (``C_ng_ml``, ``C_DOSE``). Each system parameter is passed through the simulation into the output (``F1``, ... ``MW``). This model has two covariates specified DOSE and WT. The initial value of these covariates is passed through as well as the values at each time point. For the DOSE this is SIMINT_CVIC_DOSE and DOSE respectively. Next secondary parameters are also provided (``kel``, ... ``kpc``). Lastly each timescale specified in the system file is also passed through with a "``ts.``" prefix:

```{r echo=FALSE}
rhandsontable(som$simout, width=600, height=300)
```

```{r warning=FALSE, fig.width=7, fig.height=3}
p = ggplot() + 
    geom_line(data=som$simout, aes(x=ts.days, y=C_ng_ml), color="blue")  +
    xlab("Time (days)")+
    ylab("C (ng/ml) (units)")
p = gg_log10_yaxis(p, ylim_min=1e3, ylim_max=3e5)
p = prepare_figure("print", p)
print(p)
```

## Simulating Population Response From IIV

Next we want to simulate the response of multiple subjects. This system has IIV specified in the following manner:
```{r warning=FALSE, comment=''}
cat(system_view(cfg, 'iiv'))
```

Different aspects of the Monte Carlo simulations can be specified. For example the following states that we want to simulate 20 subjects and the simulation is run using ```simulate_subjects```:

```{r warning=FALSE, results=FALSE}
cfg=system_set_option(cfg, group  = "stochastic",
                           option = "nsub",
                           value  = 20)

som  = simulate_subjects(parameters, cfg)
```

The output here, som, has a different structure than the output from an individual simulation. It is a list with the following elements

* ``som$subjects$parameters`` - Matrix with a row for each subject containing that subjects parameters
* ``som$subjects$secondary_parameters`` - Matrix containing the static secondary parameters (one row for each subject) 
* ``som$tcsummary`` - Data frame containing time course summary level information for the simulations there are columns for
  * Timescales (prefix ``ts.``)
  * States (prefix ``s.``) and Outputs (prefix ``o.``) with the mean (suffix ``.mean``) , median (suffix ``.median``) and upper (suffix ``.ub_ci``) and lower (suffix ``.lb_ci``) bounds on the specified confidence interval (default 95%)
* ``som$states`` - List with an element for each state. Each of these is a matrix containing the subject level predictions for that state (one row for each subject)
* ``som$output`` - List with an element for each output. Each of these is a matrix containing the subject level predictions for that output (one row for each subject)
* ``som$times`` - Data frame with a column for the simulation time (``time``) and a column for each timescale with a ``ts.`` prefix.


```{r warning=FALSE, fig.width=7, fig.height=3}
p = ggplot(som$tcsummary, aes(x=ts.days, y=o.C_ng_ml.mean)) +
           geom_ribbon(aes(ymin=o.C_ng_ml.lb_ci, 
                           ymax=o.C_ng_ml.ub_ci), 
                           fill="lightblue", 
                           alpha=0.6) +
           geom_line(linetype="solid", size=0.7, color="blue")  +
           geom_line(aes(x=ts.days, y=o.C_ng_ml.ub_ci), linetype="dashed", size=0.2, color="blue")  +
           geom_line(aes(x=ts.days, y=o.C_ng_ml.lb_ci), linetype="dashed", size=0.2, color="blue")  +
           xlab("Time (days)")+
           ylab("C (ng/ml) (units)")+
           guides(fill=FALSE) 


p     = gg_log10_yaxis(p    , ylim_min=1e3, ylim_max=3e5)
p     = prepare_figure("print", p    )
print(p)
```

## Simulating Population Response From File

Subject information can be pulled from a data file. First we need to load the data set using ``system_load_data``, here the data set is named ``SUBS``.

```{r echo=TRUE, warning=FALSE }
cfg = system_load_data(cfg, 
                       dsname    = "SUBS", 
                       data_file = system.file("ubinc", "csv", "mab_pk_subjects.csv", 
                                               package = "ubiquity"))
```
The format of the data set is shown below. There needs to be an column for the subject ID (``SIMINT_ID``) and the simulation time (``SIMINT_TIME``). Next the subject level parameters can be specified where the column headers correspond to the parameter names. If a parameter is not specified, the default value will be taken from the ```parameters``` input. In this case ``MW`` was not specified in the data set, so it will be taken from ``parameters``. 

Optionally, covariates can also be specified by name as well. For time varying covariates , multiple records can be specified for the same subject (the parameter values for that subject will remain unchanged over time). In this example ``WT`` and ``SEX`` are covariates but only ``WT`` has been defined in the system file, so ``SEX`` will be ignored. The co variate ``DOSE`` is not defined so the default from the system file will be used. 
```{r echo=FALSE}
rhandsontable(cfg$data$SUBS$values, width=800, height=300)
```

With the data set loaded, we need to link that subject file to the stochastic simulations using the data set name (``SUBS``) and then we can tell the scripts how to handle sampling from the data set:
```{r results=FALSE}
cfg=system_set_option(cfg, group  = "stochastic",
                           option = "sub_file",
                           value  = "SUBS")

cfg=system_set_option(cfg, group  = "stochastic",
                           option = "sub_file_sample",
                           value  = "with replacement")
```

Then we just use ``simulate_subjects`` to run the simulation as before:
```{r results=FALSE, warning=FALSE}
som  = simulate_subjects(parameters, cfg)
```


## Contents of ``system.txt``
```{r echo=FALSE, comment=''}
cat(readLines("system.txt"), sep = "\n")
```
