---
#title: "System Report"
#author: "Author"
#date: "1/7/2017"
output: html_document
---

```{r setup, include=FALSE, results=FALSE, warning=FALSE, echo=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
```

```{r, warning=FALSE, message=TRUE , echo=FALSE, error=TRUE, results=FALSE}
# Settign up the local variables
cfg = params$cfg
som = params$som

# Next we update the information in cfg to match that in the gui. Bascially if
# there is an overwrite specified in the cfg$gui$sysel_simulate data structure 
# we apply those here using the eval:
cfg = system_select_set(cfg, cfg$gui$sysel_simulate$pset)
cfg_option_names =  c('parameters', 
                      'options', 
                      'bolus', 
                      'covariates', 
                      'infusion_rates',
                      'iiv')

for(option_name in cfg_option_names){
  if(cfg$gui$sysel_simulate[[option_name]] != ""){
    eval(parse(text=cfg$gui$sysel_simulate[[option_name]])) }
}

# Pulling out the parameter set that was selected
pset_short = cfg$parameters$current_set
pset_long = cfg$parameters$sets[[pset_short]]$name

# Creating boolean variables to indicate the type
# of simulation that has been performed
som_individual = !(cfg$gui$check_variability)
som_population =   cfg$gui$check_variability

if(som_individual){
   # Put your individual code here
}


if(som_population){
   # Put your population code here
}

```
### Implementing reporting

**Disable the "Model Report" tab**

Edit the file model_gui.r, search for the following line and comment it out:


`cfg$gui$modelreport_files$R1$title = "Tab Title"`
`cfg$gui$modelreport_files$R1$file  = "my_report.Rmd"`


**Setting up a Customized Report**

To create a customized report copy the template file:

`r print(sprintf("library%stemplates%sr_system_report.Rmd", .Platform$file.sep, .Platform$file.sep))`

Into the main directory and name it ` my_report.Rmd`. Then edit the
model_gui.r file and point it to the new report file:

`cfg$gui$modelreport_files$R1$title = "Tab Title"`
`cfg$gui$modelreport_files$R1$file  = "my_report.Rmd"`

Next you can customize that file using the information below.

**Customizing the report**

To customize this report, you have access the following variables:

 * cfg - System configuration variable
 * som - Mapped simulation output
 * parameters - current system parameters
 
To understand the structure of som and access the components of cfg see the documentation at: 

[presentation.ubiquity.grok.tv](http://presentation.ubiquity.grok.tv)

Go to the R-Workflow part and look at the output of run_simulation_ubiquity
for individual simulations and simulate_subjects for the population
simulations.

`r if(!som_individual){"\\begin{comment}"}`
Simulation Type: **Individual**

This text is shown when a individual simulation has been performed. To make
sure that R-code is only executed for individual simulation output it is
necessary to wrap that code in a conditional statement:

`r if(!som_individual){"\\end{comment}"}`



`r if(!som_population){"\\begin{comment}"}`
Simulation Type: **Population**

This text is shown when a population simulation has been performed. To make
sure that R-code is only executed for population simulation output it is
necessary to wrap that code in a conditional statement:

`r if(!som_population){"\\end{comment}"}`

```{r fig.width=7, fig.height=4}

# uncomment the text below and replace OUTPUT with the model output
# and replace TS with the actual timescale
#p = ggplot() 
#p = p + geom_line(data=som$simout, aes(x=TS, y=OUTPUT  , color='OUTPUT'), linetype='solid', size=0.9)
#p = p + xlab('Time (TIME UNITS)')                     
#p = p + ylab('Output')                         
#p = prepare_figure('present', p)              
#p = p + scale_colour_manual(values=c("OUTPUT"="darkblue"))   
#p = p + theme(legend.title = element_blank()) 
#p = p + theme(legend.position = 'bottom')
#p
```

```{r results = "asis"}

kable(t(parameters), col.names=c(''), caption="System Parameters")

```


